Turbo fast no compromise gravitational wave parameter estimation

### 2023-01-12 src/tex/ms.tex
Add citation to heterodyned likelihood section
Add citation

Since the computational cost of evaluating a waveform model scales linearly with
the number of sample points either in the time or frequency domain, the
computational burden for longer-duration signals is often quite large. To
reduce the computational cost, there are a number of methods to reduce the
number of basis points one would need to compute the likelihood faithfully
\kw{cite}. We use heterodyned likelihood \kw{cite} (also named relative binning
in \kw{cite}) in this work. The idea behind heterodyned likelihood can be
summarized as the following: the integrand in eq. \ref{eq: innerproduct} is
a highly oscillatory function, so one has to sample the integrand with
sufficiently dense sampling to compute the integral faithfully. The number of
sample points needed would be much smaller if the integrand was smooth. Given a
pair of waveform parameters $\mathbf{\theta}$ and $\mathbf{\theta_0}$ that are
close to each other, the waveforms generated using the pair of parameters are
similar to each other, this means the ratio between the waveforms is a smoothly
varying function. Given a reference waveform $h(\mathbf{\theta_0})$, we can
exploit this similarity between waveforms to reduce the number of sample points
needed to compute the likelihood for the set of $\mathbf{\theta}$ that is
similar to $\mathbf{\theta_0}$. We decompose the integrand into two parts: a
highly oscillatory part that depends only on the reference waveform and the
data, hence it only needs to be evaluated once; and a smoothly varying part that
depends on the target waveform parameters $\theta$, that needs to be evaluated
for every new likelihood evaluation. Because the part that depends on the target
waveform parameters is smooth, we can use far fewer sample points to compute the
integral with sufficient accuracy.

@@

In MCMC, the posterior distribution is approximated by a Markov chain that
converges to the target distribution. The Markov chain is constructed by
iteratively proposing a new point in the parameter space based on the current
location of the chain. The proposed point is accepted with a probability that is
usually set to be proportional to the ratio of the posterior density evaluated
at the proposed point and the current point. The chain can either accept the
proposal and move to the new location or reject the proposal and stay at the
current location. This process is repeated until the chain converges to the
target distribution. The samples generated by the chain are then used to
estimate the quantities of interest, such as the mean and credible intervals of
the source parameters. In practice, since we do not know the target distribution
ahead of time, the MCMC process is usually repeated until a certain criterion is
met, such as a Gelman-Rubin convergence statistic \cite{Gelman-rhat} lower than
a threshold or simply after a certain number of iterations.

@@

Transition kernels that leverage gradient information of the target
distribution can help to address this issue of shortening steps in a high
dimensional space. Instead of proposing a new point by drawing from a Gaussian,
one can use the gradient evaluation at the current location to propose a new
point. For example, Metropolis-adjusted Langevin algorithm (MALA) \kw{cite}
place a unit Gaussian at the tip of the gradient vector at the current position,
\begin{align}
    \mathbf{x} = \mathbf{x_0} + \tau \nabla\log{p(\mathbf{x_0})} + \sqrt{2\tau}N(0,\mathbf{I}),
\end{align}
where $\tau$ is the step size, which is a tuning parameter. Compared to a
Gaussian centers at the current location, the MALA transition kernel is more
likely to propose a point into the higher posterior density region because of
the gradient term, which helps to boost the acceptance rate.

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to be superior in terms of
performance when compared to other gradient-free algorithms in many practical
examples\kw{cite}, there are still a number of classes of problems
gradient-based samplers do not solve well. \footnote{To be specific, we are
referring to sampling algorithms that use first-order derivatives here. Sampling
algorithms that use the information of higher order derivatives such as manifold-MALA
and Riemannian-HMC can in principle decorrelate local correlation in the target
distribution, however, they often have instability issues when they are used on
real-life applications, so they are not used often in practice.} For example,
target distribution that exhibits local correlation is hard to deal with,
since by construction first order gradient-based algorithms can only handle
global correlation structure. Another example is multi-modality. If there are
multiple modes in the target distribution, an independent chain will likely 
be trapped in one mode and take an extremely long time to transverse between the
modes. This means the relative weights between modes will much longer to sample
compared to the shape of each mode.

>>>>

Since the computational cost of evaluating a waveform model scales linearly with
the number of sample points either in the time or frequency domain, the
computational burden for longer-duration signals is often quite large. To reduce
the computational cost, there are a number of methods to reduce the number of
basis points one would need to compute the likelihood faithfully
\cite{Field:2011mf, Field:2013cfa, Smith:2016qas, Vinciguerra:2017ngf}. We use
heterodyned likelihood \cite{Cornish:2021lje} (also named relative binning in
\cite{Zackay:2018qdy}) in this work. The idea behind heterodyned likelihood can
be summarized as the following: the integrand in eq. \ref{eq: innerproduct} is a
highly oscillatory function, so one has to sample the integrand with
sufficiently dense sampling to compute the integral faithfully. The number of
sample points needed would be much smaller if the integrand was smooth. Given a
pair of waveform parameters $\mathbf{\theta}$ and $\mathbf{\theta_0}$ that are
close to each other, the waveforms generated using the pair of parameters are
similar to each other, this means the ratio between the waveforms is a smoothly
varying function. Given a reference waveform $h(\mathbf{\theta_0})$, we can
exploit this similarity between waveforms to reduce the number of sample points
needed to compute the likelihood for the set of $\mathbf{\theta}$ that is
similar to $\mathbf{\theta_0}$. We decompose the integrand into two parts: a
highly oscillatory part that depends only on the reference waveform and the
data, hence it only needs to be evaluated once; and a smoothly varying part that
depends on the target waveform parameters $\theta$, that needs to be evaluated
for every new likelihood evaluation. Because the part that depends on the target
waveform parameters is smooth, we can use far fewer sample points to compute the
integral with sufficient accuracy.

@@

In MCMC, the posterior distribution is approximated by a Markov chain that
converges to the target distribution. The Markov chain is constructed by
iteratively proposing a new point in the parameter space based on the current
location of the chain. The proposed point is accepted with a probability that is
usually set to be proportional to the ratio of the posterior density evaluated
at the proposed point and the current point. The chain can either accept the
proposal and move to the new location or reject the proposal and stay at the
current location. This process is repeated until the chain converges to the
target distribution. The samples generated by the chain are then used to
estimate the quantities of interest, such as the mean and credible intervals of
the source parameters. In practice, since we do not know the target distribution
ahead of time, the MCMC process is usually repeated until a certain criterion is
met, such as a Gelman-Rubin convergence statistic \cite{10.2307/2246093} lower than
a threshold or simply after a certain number of iterations.

@@

Transition kernels that leverage gradient information of the target
distribution can help to address this issue of shortening steps in a high
dimensional space. Instead of proposing a new point by drawing from a Gaussian,
one can use the gradient evaluation at the current location to propose a new
point. For example, Metropolis-adjusted Langevin algorithm (MALA) \cite{10.2307/2346184}
place a unit Gaussian at the tip of the gradient vector at the current position,
\begin{align}
    \mathbf{x} = \mathbf{x_0} + \tau \nabla\log{p(\mathbf{x_0})} + \sqrt{2\tau}N(0,\mathbf{I}),
\end{align}
where $\tau$ is the step size, which is a tuning parameter. Compared to a
Gaussian centers at the current location, the MALA transition kernel is more
likely to propose a point into the higher posterior density region because of
the gradient term, which helps to boost the acceptance rate.

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to be superior in terms of
performance when compared to other gradient-free algorithms in many practical
examples , there are still a number of classes of problems
gradient-based samplers do not solve well. \footnote{To be specific, we are
referring to sampling algorithms that use first-order derivatives here. Sampling
algorithms that use the information of higher order derivatives such as manifold-MALA
and Riemannian-HMC \cite{RMHMC} can in principle decorrelate local correlation in the target
distribution, however, they often have instability issues when they are used on
real-life applications, so they are not used often in practice.} For example,
target distribution that exhibits local correlation is hard to deal with,
since by construction first order gradient-based algorithms can only handle
global correlation structure. Another example is multi-modality. If there are
multiple modes in the target distribution, an independent chain will likely 
be trapped in one mode and take an extremely long time to transverse between the
modes. This means the relative weights between modes will much longer to sample
compared to the shape of each mode.
